\chapter{Parsing Expression Grammars}\label{peg}

I'm working on a new parser combinator library for Factor based on
Parsing Expression Grammars and Packrat parsers. This is based on what
I learnt from writing a packrat parser in Javascript.

It's progressing quite well and already fixes some problems in the
existing parser combinator library I wrote. The main issue with that
one is it's not tail recursive and some combinations of parsers can
run out of call stack.

The new library is in the \texttt{peg} vocabulary. I haven't yet implemented
the packrat side of things though so it is slow on large grammars and
inputs.

I've also done a proof of concept of something I've been meaning to do
for awhile. That is writing a parser for EBNF (or similar) that
produces Factor code in the form of parser combinators to implement
the described grammar. The code for this is in the \texttt{peg.ebnf}
vocabulary. It allows you to embed an EBNF-like language and have
Factor words generated for each rule:
\begin{alltt}
<EBNF
digit  = '1' | '2' | '3' | '4' .
number = digit { digit } .
expr   = number {('+' | '-' ) number } .
EBNF>
\end{alltt}
This example would create three Factor words. \texttt{digit}, \texttt{number} and
\texttt{expr}. These words return parsers that can be used as normal:
\begin{alltt}
"123" number parse 
"1" digit parse
"1+243+342" expr parse
\end{alltt}
The EBNF accepted allows for choice, zero or more repetition, optional
(exactly 0 or 1), and grouping. The generated AST is pretty ugly so by
default it works best as a syntax checker. You can modify the
generated AST with action productions:
\begin{alltt}
<EBNF
digit  = '1' | '2' | '3' | '4' => convert-to-digit .
number = digit { digit }       => convert-to-number .
expr   = number '+' number     => convert-to-expr .
EBNF>
\end{alltt}
An action is a factor word after the \texttt{=>}. The word receives the AST
produced from the rule on the stack and it can replace that with a new
value that will be used in the AST. So \texttt{convert-to-expr} above might
produce a tuple holding the expression values (by default, a sequence
of terms in the rule are stored in a vector):
\begin{alltt}
TUPLE: ast-expr lhs operator rhs ;
C: <ast-expr> ast-expr
: convert-to-expr ( old -- new )
  first3 <ast-expr> ;
\end{alltt}
The generated code is currently pretty ugly, mainly due to it being a
quick proof of concept. I'll try doing a few grammars and tidy it up,
changing the interface if needed, as I go along.

As an experiment I did a grammar for the PL/0 programming
language. It's in \texttt{peg.pl0}. The grammar from the wikipedia article
is:
\begin{alltt}
program = block "." .

block = [ "const" ident "=" number {"," ident "=" number} ";"]
        [ "var" ident {"," ident} ";"]
        { "procedure" ident ";" block ";" } statement .

statement = [ ident ":=" expression | "call" ident |
            "begin" statement {";" statement } "end" |
            "if" condition "then" statement |
            "while" condition "do" statement ].

condition = "odd" expression |
            expression ("="|"#"|"<"|"<="|">"|">=") expression .

expression = [ "+"|"-"] term { ("+"|"-") term}.

term = factor {("*"|"/") factor}.

factor = ident | number | "(" expression ")".
\end{alltt}
The Factor grammar is very similar:
\begin{verbatim}
: ident ( -- parser )
  CHAR: a CHAR: z range 
  CHAR: A CHAR: Z range 2array choice repeat1 
  [ >string ] action ;

: number ( -- parser )
  CHAR: 0 CHAR: 9 range repeat1 [ string>number ] action ;

<EBNF
program = block '.' .
block = [ 'const' ident '=' number { ',' ident '=' number } ';' ]
        [ 'var' ident { ',' ident } ';' ]
        { 'procedure' ident ';' [ block ';' ] } statement .
statement = [ ident ':=' expression | 'call' ident |
              'begin' statement {';' statement } 'end' |
              'if' condition 'then' statement |
              'while' condition 'do' statement ] .
condition = 'odd' expression |
            expression ('=' | '#' | '<=' | '<' | '>=' | '>') expression .
expression = ['+' | '-'] term {('+' | '-') term } .
term = factor {('*' | '/') factor } .
factor = ident | number | '(' expression ')'
EBNF>
\end{verbatim}
This grammar as defined works and can parse PL/0 programs. I'll extend
this as I improve the EBNF routines, adding actions, etc to generated
a decent AST.

\section{Parsing Arithmetic Expressions}

This is it without parenthesis handling. I'll leave it as an exercise
to add that. I included code to compile the ast into a factor
expression. Use like:

\begin{verbatim}
"2+3*4-6/2" expr parse parse-result-ast .

"2+3*4-6/2" expr parse parse-result-ast ast-compile dup .

"2+3*4-6/2" expr parse parse-result-ast ast-compile call
\end{verbatim}

The grammar is a direct translation of the one shown in the wikipedia
peg article:

\begin{verbatim}
http://en.wikipedia.org/wiki/Parsing_expression_grammar
\end{verbatim}

The only complication is transforming the result of using 'repeat0'
into a tree by doing a fold. This is standard practice with parsing
expression grammars that don't handle left recursion. When I add left
recursion this won't be needed and the grammar will be simplified.
See this article for details:

\begin{verbatim}
http://vpri.org/pdf/packrat_TR-2007-002.pdf
\end{verbatim}

\begin{verbatim}
USING: sequences kernel math words math.parser namespaces peg ;
IN: scratchpad

: seq* ( quot -- parser )
 { } make seq ; inline

: choice* ( quot -- parser )
 { } make choice ; inline

TUPLE: operator lhs op rhs  ;
C: <operator> operator

: operator-fold ( lhs seq -- value )
 #! Perform a fold of a lhs, followed by a sequence of pairs being
 #! { operator rhs } in to a tree structure of the correct precedence.
 swap [ first2 <operator> ] reduce ;

: digits ( -- parser )
 CHAR: 0 CHAR: 9 range repeat1 [ string>number ] action ;

: plus ( -- parser )
 "+" token ;

: minus ( -- parser )
 "-" token ;

: multiply ( -- parser )
 "*" token ;

: divide ( -- parser )
 "/" token ;

DEFER: expr

: value ( -- parser )
 [ digits , [ expr ] delay , ] choice* ;

: product ( -- parser )
 [
   value ,
   [
     [ multiply , divide , ] choice* ,
     value ,
   ] seq* repeat0 ,
 ] seq* [ first2 operator-fold ] action ;

: sum ( -- parser )
 [
   product ,
   [
     [ plus , minus , ] choice* ,
     product ,
   ] seq* repeat0 ,
 ] seq* [ first2 operator-fold ] action ;

: expr ( -- parser )
 sum ;

GENERIC: (compile) ( ast -- )

M: number (compile) ( ast -- )
 , ;

M: operator (compile) ( ast -- )
 dup operator-lhs (compile)
 dup operator-rhs (compile)
 operator-op "math" lookup , ;

: ast-compile ( ast -- quot )
 [ (compile) ] [ ] make ;
\end{verbatim}
